Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-12 doc/invt.tex
add dinvacmp.gp

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}
along $z$ over a wide range.
%
This alleviates the above problem of local trapping
as the system is now forced to travel more flexibly
along the $z$ direction,
which often represents
a direction of slow motion.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential must exactly cancel the PMF
to achieve a flat distribution.

@@

Naturally, the updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}
determines the precision of the final bias potential,
hence that of PMF.
%
For the WL algorithm, it is found
that asymptotically the optimal magnitude of updating $\ln f$
should be given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' is defined as
the number of simulation steps
divided by the number of bins.

In this study,
we shall reexamine the inverse-time formula,
and similarly optimize schedules
for updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
It turns out that the optimal updating schedule
can be sensitive the simulation length.
%
We also shall show that
the single-bin scheme used by the WL algorithm
is best in asymptotic convergence.
%
For finite-length simulations, however,
metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

In this section,
we first review the basics and fix notations
in Sec. \ref{sec:background}.
%
We then prove the optimality
of the inverse-time formula
for the single-bin WL algorithm
in Sec. \ref{sec:single-bin}.
%
Next, we discuss the generalized
multiple-bin schemes
in Secs. \ref{sec:multiple-bin}
and \ref{sec:band-matrix}.
%

@@

\subsubsection{$1/t$ formula}

@@

@@

As shown above, with fixed endpoint value of $q(t)$,
the optimization of the error is equivalent to
that of the asymptotic error.
%
We now rewrite Eq. \eqref{eq:error_asym1} as
%
\begin{equation}
  \Err_A
  =
  \int_0^t
    {\mathcal L} \bigl[ q(\tau)\bigr]
    \, d\tau,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(\tau) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl( q(\tau) \bigr)
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl( q(\tau) \bigr)
  \; \dot q^2( \tau ).
  \label{eq:error_asym_lagr}
\end{align}
%
To minimize the functional,
we notice that the Lagrangian given by Eq. \eqref{eq:error_asym_lagr}
does not explicitly depend on time, $t$.
%
Thus, the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved, and in this case,
the Hamiltonian and the Lagrangian coincide.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(\tau) \bigr]
  \;
  \dot q^2(\tau)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating integrate the square root of the left-hand side,
%
\begin{equation}
  \int_0^{ q(\tau) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, d\tau
  ,
  \label{eq:q_opt}
\end{equation}
%
we get an implicit equation for the optimal $\alpha(\tau)$
via its integral $q(\tau)$.

@@

\subsubsection{Variationally optimal schedule, $\alpha(t)$}

@@

\subsubsection{\label{sec:optlambda}
Optimal parameter $\lambda$}

@@

We now consider a class of commonly-used updating scheme
for sampling a uniform distribution, $p_i = 1/n$.
%
These schemes are homogeneous,
or translational invariant with respect to $z$.
%
This invariance is natural for a periodic variable\cite{dama2014}.
%
Below, however, we focus on a $z$ that is
under the reflective boundary conditions.

@@

In the above, we have assumed the inverse-time schedule,
Eq. \eqref{eq:alpha_invtlambda}.
%
In Fig. \ref{fig:alpha_cmp},
we compared the inverse-time schedule
with the optimal $\lambda$,
and the optimal schedule from integrating
Eq. \eqref{eq:q_opt}.
%
The difference was larger for the global sampling scheme.
%
\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/alphacmp.pdf}
  }
  \caption{
    \label{fig:alpha_cmp}
    Optimal schedules, $\alpha(t)$, from Eq. \eqref{eq:q_opt},
    versus the inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal parameter, $\lambda$.
    %
    For fair comparison,
    the $q(t) = \int_0^t \alpha(\tau) \, d\tau$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}

>>>>

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}
along $z$.
%
This alleviates the above problem of local trapping
as the system is now forced to travel more flexibly
along the $z$ direction,
which often represents
a direction of slow motion.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential must exactly cancel the PMF
to achieve a flat distribution.

@@

Naturally, the updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}
determines the precision of the final bias potential,
hence that of PMF.
%
For the WL algorithm, it is found
that asymptotically the optimal magnitude of updating $\ln f$
should be given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' is defined as
the number of simulation steps
divided by the number of bins.
%
For metadynamics, ...

In this study,
we shall reexamine the inverse-time formula,
and similarly optimize schedules
for updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
It turns out that the optimal updating schedule
can be sensitive the simulation length.
%
We also shall show that
the single-bin updating scheme used by the WL algorithm
is best in asymptotic convergence.
%
For finite-length simulations, however,
metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

In this section,
we first review the basics and fix notations
in Sec. \ref{sec:background}.
%
We then prove the optimality
of the inverse-time formula
for the single-bin updating WL algorithm
in Sec. \ref{sec:single-bin}.
%
Next, we discuss the generalized
multiple-bin schemes
in Secs. \ref{sec:multiple-bin}
and \ref{sec:band-matrix}.
%

@@

\subsubsection{Updating magnitude and the inverse-time formula}

@@

\subsection{\label{sec:mbin_opta}
Optimal schedule for the multiple-bin updating scheme}

@@

As shown above, with fixed endpoint value of $q(t)$,
the optimization of the error is equivalent to
that of the asymptotic error.
%
We now rewrite Eq. \eqref{eq:error_asym1} as
%
\begin{equation}
  \Err_A
  =
  \int_0^t
    {\mathcal L} \bigl[ q(\tau)\bigr]
    \, d\tau,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(\tau) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl( q(\tau) \bigr)
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl( q(\tau) \bigr)
  \; \dot q^2( \tau ).
  \label{eq:error_asym_lagr}
\end{align}
%
To minimize the functional,
we notice that the Lagrangian given by Eq. \eqref{eq:error_asym_lagr}
does not explicitly depend on time, $t$.
%
Thus, the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved, and in this case,
the Hamiltonian and the Lagrangian coincide.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(\tau) \bigr]
  \;
  \dot q^2(\tau)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_0^{ q(\tau) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, d\tau
  ,
  \label{eq:q_opt}
\end{equation}
%
we get an implicit equation for the optimal $\alpha(\tau)$
via its integral $q(\tau)$.

@@

\subsubsection{Inverse-time schedule}

@@

\subsubsection{\label{sec:optlambda}
Optimal proportionality constant}

@@

We now consider a class of updating schemes
that are homogeneous,
or translationally invariant. % with respect to $z$,
%
As we shall see,
these schemes include the Gaussian updating scheme
commonly used in metadynamics.
%
Although the translational invariance is
natural for a periodic variable\cite{dama2014},
it can be also extended to an open-end variable
by imposing the reflective boundary conditions.
%
For simplicity, we shall assume that our target
distribution is uniform, or $p_i = 1/n$, below.

@@

In the above, we have assumed the inverse-time schedule,
Eq. \eqref{eq:alpha_invtlambda}.
%
In Fig. \ref{fig:alpha_cmp},
we compared the inverse-time schedule
with the optimal $\lambda$,
and the optimal schedule from integrating
Eq. \eqref{eq:q_opt}.
%
The difference was larger for the global sampling scheme.
%
Compare with Eqs. \eqref{eq:alpha_limit1} and \eqref{eq:alpha_limit2}.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/dinvacmp.pdf}
  }
  \caption{
    \label{fig:alpha_cmp}
    Optimal schedules, in terms of $\dot \alpha(t)$, from Eq. \eqref{eq:q_opt},
    versus the inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal parameter, $\lambda$.
    %
    For fair comparison,
    the $q(t) = \int_0^t \alpha(\tau) \, d\tau$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}
